---
title: "Final Code"
author: Natasha, Sophia, Madison
output:
  pdf_document: default
  html_notebook: default
---
# _Setting up the dataset:_
```{r}
library(tidyverse)
pol <- read_csv('Pol_Library_Visits_Dataset.csv')
```

## _Filtering the data:_

- The total pollinator dataset was filtered so that only Hymenoptera families with more than 50 samples were included in the dataset. These families are Andrenidae, Apidae, and Halictidae. Of those families, only the largest genera with greater than 20 samples, Apis, Bombus, Lassioglossum, Melissodes, Perdita were included. After all the data was filtered, more than 800 observations remained.
- We are using diveristy as the statistic for measuring population dynamics of the pollinator species. Diversity is calculated per unique GPS location. It is assumed that there was uniform sampling of values. 

```{r}
pol.hymenoptera <- pol%>% 
  filter(order == 'Hymenoptera') %>% 
  filter(length(family) >= 50 ) %>% 
  group_by(genus) %>% 
  filter(!is.na(genus)) %>% 
  filter(length(genus) >= 20)

simpsons_diversity_index <-  function (vector){
    N <- sum(vector)
    p <- sum(((vector)/sum(vector))^2)
    print(1-p)
}

pol.hymenoptera<- pol.hymenoptera%>% 
  group_by(long, lat) %>% 
  mutate(n = n()) %>% 
  mutate(diversity = simpsons_diversity_index(n))

```

# _Question 1:_
_Author_: Sophia Fan
_Overview of Analysis_ :

Necessary packages
```{r}
library(plyr)
library(broom)
library(lmerTest)
library(lme4)
library(MuMIn)
library(car)
```

Data is scaled to allow for unskewed analysis through models
```{r}
pol.hymen.scaled <- pol.hymenoptera

pol.hymen.scaled$elevation <- scale(pol.hymen.scaled$elevation, center = TRUE, scale = TRUE)
pol.hymen.scaled$lat <- scale(pol.hymen.scaled$lat, center = TRUE, scale = TRUE)
pol.hymen.scaled$long <- scale(pol.hymen.scaled$long, center = TRUE, scale = TRUE)
pol.hymen.scaled$avws <- scale(pol.hymen.scaled$avws, center = TRUE, scale = TRUE)
pol.hymen.scaled$relhum <- scale(pol.hymen.scaled$relhum, center = TRUE, scale = TRUE)
pol.hymen.scaled$airtemp <- scale(pol.hymen.scaled$airtemp, center = TRUE, scale = TRUE)
```


Per GPS point, diversity index
```{r}
#Checking for covariance
initial.mod <- lm(diversity ~ elevation + avws + airtemp + relhum + lat, data = pol.hymen.scaled)
summary(initial.mod)
vif(initial.mod)
# elevation and latitude are correlated ...redundant info!
# most significant factors are relhum, airtemp, and elevation

initial.mod.ele <- lm(diversity ~ elevation + avws + airtemp + relhum, data = pol.hymen.scaled)
initial.mod.lat <- lm(diversity ~ lat + avws + airtemp + relhum, data = pol.hymen.scaled)

summary(initial.mod.ele)
summary(initial.mod.lat)

# R^2 value 0.03142 for initial.mod.lat, better than initial.mod.ele (0.02539) --> only use lat

#Optimize random effects
mod_intonly <- lmer(diversity ~ lat*relhum*airtemp + (1|county), REML = TRUE, data = pol.hymen.scaled)

mod_intslope <- lmer(diversity ~ lat*relhum*airtemp + (1+lat|county), REML = TRUE, data = pol.hymen.scaled)

AICc(mod_intonly, mod_intslope)
#mod_intslope has the smaller AIC value
```

```{r}
#Optimize fixed effects
mod_full <- lmer(diversity ~ lat*relhum*airtemp + (1 + lat|county), REML = FALSE, data = pol.hymen.scaled)

mod_lat <- lmer(diversity ~ lat + (1 + lat|county), REML = FALSE, data = pol.hymen.scaled)

mod_relhum <- lmer(diversity ~ relhum + (1 + lat|county), REML = FALSE, data = pol.hymen.scaled)

mod_airtemp <- lmer(diversity ~ airtemp + (1 + lat|county), REML = FALSE, data = pol.hymen.scaled)

mod_lathum <- lmer(diversity ~ lat*relhum + (1 + lat|county), REML = FALSE, data = pol.hymen.scaled)

mod_lattemp <- lmer(diversity ~ lat*airtemp + (1 + lat|county), REML = FALSE, data = pol.hymen.scaled)

mod_humtemp <- lmer(diversity ~ relhum*airtemp + (1 + lat|county), REML = FALSE, data = pol.hymen.scaled)

mod_lathum_nointer <- lmer(diversity ~ lat + relhum + (1 + lat|county), REML = FALSE, data = pol.hymen.scaled)

mod_lattemp_nointer <- lmer(diversity ~ lat + airtemp + (1 + lat|county), REML = FALSE, data = pol.hymen.scaled)

mod_humtemp_nointer <- lmer(diversity ~ relhum + airtemp + (1 + lat|county), REML = FALSE, data = pol.hymen.scaled)

mod_nofix <- lmer(diversity ~ 1 + (1 + lat|county), REML = FALSE, data = pol.hymen.scaled)

AICc(mod_full,
     mod_lat,
     mod_relhum,
     mod_airtemp,
     mod_lathum,
     mod_lattemp,
     mod_humtemp,
     mod_lathum_nointer,
     mod_lattemp_nointer,
     mod_humtemp_nointer,
     mod_nofix)

#mod_full was the best model

summary(mod_full)
#relhum and lat:airtemp were significant
```
## *things to add*: 
maybe some plots showing diveristy and the environmental effects. A comprehensive dataset that shows all the model outputs( I am still working on this myself.)

# _Question 3_: How does human population affects pollinator diversity and distribution. 
_Author_: Natasha Dhamrait
_Overview of analysis_: GPS mapping, spatial stats, raster data

Necessary packages:
```{r}
library(MASS)
library(PerformanceAnalytics)
library(nlme)
library(sp)
library(ape)
library(rgdal)
library(raster)
library(maps)
library(viridis)
library(reshape2)
```


First need to convert all the GPS coordinates to workable coordinates
```{r}
coord.pol <- SpatialPoints(cbind(pol.hymenoptera$long, pol.hymenoptera$lat),
                                  proj4string=CRS("+proj=longlat +ellps=WGS84"))

sinu.pol <- spTransform(coord.pol, CRS("+proj=sinu +ellps:7271"))

pol.hymenoptera <- bind_cols(pol.hymenoptera, as.data.frame(sinu.pol))
```

Mapping for the USA
```{r}
# all of the USA
map.usa <- map_data("world", region="usa")
 
# map with state boundaries only including the continuous states
us_states <- map_data("state")

ggplot() + 
  geom_polygon(data=us_states, aes(x=long, y=lat, group=group), colour="black", fill = "grey32") + 
  coord_fixed()+
  labs(x = "Longitude", y = "Latitude", title = "Distribution of Observations in the Hymenoptera Order")+
  geom_point(data= pol.hymenoptera, aes(x=long, y=lat, colour= family), size = 3 , alpha = 0.2) +
    theme_classic()

```


plotting graphs for each of the three states and colouring by family/order. To plot each of the 3 states seperately, filter the us_states map data. to seperate per state, additionally filter the pol.hymenoptera df by state. limit the x and y axis so only shows the map area with data points. 
```{r}
# Texas only
texas_only <- us_states %>% 
  filter(region == "texas")   

ggplot() + 
  geom_polygon(data=texas_only, aes(x=long, y=lat, group=group), colour="black", fill = "grey32") + 
  coord_fixed() + xlim(-107,-93)+ ylim(25,37)+
  labs(x = "Longitude", y = "Latitude", title = "a)")+
  geom_point(data= pol.hymenoptera, aes(x=long, y=lat, colour= family), size = 3 , alpha = 0.2) +
    theme_classic()

# North Dakota only
nd_only <- us_states %>% 
  filter(region == "north dakota")   

pol.hymen.nd <- pol.hymenoptera %>% 
  filter(state == "North Dakota")

ggplot() + 
  geom_polygon(data=nd_only, aes(x=long, y=lat, group=group), colour="black", fill = "grey32") + 
  coord_fixed()+
  labs(x = "Longitude", y = "Latitude", title = "b)")+
  geom_point(data= pol.hymen.nd, aes(x=long, y=lat, colour= genus), size = 3.5 , alpha = 0.2, show.legend = FALSE) +
    theme_classic()+
  guides(alpha = FALSE)

# Colorodo only
col_only <- us_states %>% 
  filter(region == "colorado")   

ggplot() + 
  geom_polygon(data=col_only, aes(x=long, y=lat, group=group), colour="black", fill = "grey32") + 
  coord_fixed() + xlim(-110,-101)+ ylim(36.5,41.5)+ 
  labs(x = "Longitude", y = "Latitude", title = "c)")+
  geom_point(data= pol.hymenoptera, aes(x=long, y=lat, colour= family), size = 3 , alpha = 0.2, show.legend = FALSE) +
    theme_classic()
```

Overlay of pollinator points over a map of human population density. 
```{r}
#download a human populaiton raster from online, assign, extract points for later...
pop <- raster("w001001.adf") 
extractions.pop<- raster::extract(pop, coord.pol, df = TRUE, method="bilinear")
colnames(extractions.pop)[2] <- "hum.pop"

pol.hymenoptera <- as.data.frame(bind_cols(pol.hymenoptera, extractions.pop)) # Note order of rows!

#making raster points into a dataframe so i can plot in ggplot and make nice colors
raster_points = rasterToPoints(pop)
raster_df = data.frame(raster_points)
```


Plotting with ggplot the raster and points, per state, zommed in. To plot each state, fisrt you need to filter the raster data so the scale is not skewed by big cities, like New York. Then plot the map, raster, and points. modify the axis and color/labs to best illustrate the points.
```{r}

#texas 
raster_df.filtered.tx <- raster_df %>% 
  filter(x <= -100) %>% 
  filter(x >= -104) %>% 
  filter(y >= 32.5) %>% 
  filter(y <= 36) %>% 
  rename(h.density = w001001)

ggplot() +
  geom_polygon(data=texas_only, aes(x=long, y=lat, group=group), colour="black", fill = "grey32") + 
  geom_tile(data=raster_df.filtered.tx, aes(x=x,y=y,fill=h.density))+
  scale_fill_gradient(low = "grey25", high = "grey80", "Human Population")+
  coord_fixed()+ xlim(-103.5,-100.5)+ ylim(33,35.5)+
  geom_point(data= pol.hymenoptera, aes(x=long, y=lat, colour = diversity), size = 3 , alpha = 0.5)+
  scale_colour_viridis("Pollinator Diveristy")+
  labs(x = "Longitude", y = "Latitude", title = "a)")+
  theme_classic()

#north dakota
raster_df.filtered.nd <- raster_df %>% 
  filter(x <= -96) %>% 
  filter(x >= -100) %>% 
  filter(y >= 45.5) %>% 
  filter(y <= 48.5) %>% 
  rename(h.density = w001001)

ggplot() +
  geom_polygon(data=nd_only, aes(x=long, y=lat, group=group), colour="black", fill = "grey32") + 
  geom_tile(data=raster_df.filtered.nd, aes(x=x,y=y,fill=h.density), show.legend = FALSE)+
  scale_fill_gradient(low = "grey25", high = "grey80")+
  coord_fixed()+ xlim(-99.8,-96.7)+ ylim(45.8,48.3)+
  geom_point(data= pol.hymenoptera, aes(x=long, y=lat, colour = diversity), size = 3 , alpha = 0.5, show.legend = FALSE)+
  scale_colour_viridis()+
  geom_jitter()+
  labs(x = "Longitude", y = "Latitude", title = "b)")+
  theme_classic()

#colorado
raster_df.filtered.nd <- raster_df %>% 
  filter(x <= -100) %>% 
  filter(x >= -110) %>% 
  filter(y >= 36) %>% 
  filter(y <= 43) %>% 
  rename(h.density = w001001)

ggplot() +
  geom_polygon(data=col_only, aes(x=long, y=lat, group=group), colour="black", fill = "grey32") + 
  geom_tile(data=raster_df.filtered.nd, aes(x=x,y=y,fill=h.density), show.legend = FALSE)+
  scale_fill_gradient(low = "grey25", high = "grey80")+
  coord_fixed()+ xlim(-107.5,-100)+ ylim(38,43)+
  geom_point(data= pol.hymenoptera, aes(x=long, y=lat, colour = diversity), size = 3 , alpha = 0.5, show.legend = FALSE)+
  scale_colour_viridis()+
  geom_jitter()+
  labs(x = "Longitude", y = "Latitude", title = "c)")+
  theme(legend.title = element_blank())+
  theme_classic()
```

*Spatial stats*: MoranIs, the different projection models. 

For the models, spacial difference is included as a random effect, allowing for closer sites to be more similar than other sites. 

  the gls function, used to inclue spatial variation as a random effect in linear models, cannot work if there are differences between gps coordinates that are zero distances apart (replicates of data at the same gps point). For these models, since we are only interested in diversity and human population per gps coordinate, it is okay to only use one GPS replicate per location. 
```{r}
pol.hymen_one.replicate <- pol.hymenoptera %>% 
  dplyr::select(coords.x1, coords.x2, hum.pop, diversity, long, lat) %>% 
    group_by(hum.pop, long, lat) %>% 
  distinct(coords.x1, coords.x2, diversity) 
```

Basic tests: How does human population relate to pollinator diveristy
```{r}
cor.test(pol.hymen_one.replicate$diversity, pol.hymen_one.replicate$hum.pop)
```
There is very low variation. Let's see if this improves when including spatial variation as a random effect. 


Projection models: Using AIC to test which model is the best model
```{r}
mod <- lm(diversity ~ hum.pop, data = pol.hymen_one.replicate)

mod.exp <- gls(diversity ~ hum.pop, data=pol.hymen_one.replicate,
               corr= corSpatial(form= ~coords.x1+coords.x2, type ="exponential"))

mod.gau <- gls(diversity ~ hum.pop, data=pol.hymen_one.replicate, 
               corr=corSpatial(form= ~coords.x1+coords.x2, type ="gaussian"))

mod.lin <-gls(diversity ~ hum.pop, data=pol.hymen_one.replicate, 
               corr=corSpatial(form= ~coords.x1+coords.x2, type ="linear"))

mod.rat <- gls(diversity ~ hum.pop, data=pol.hymen_one.replicate, 
               corr=corSpatial(form= ~coords.x1+coords.x2, type ="rational"))

mod.sp <- gls(diversity ~ hum.pop, data=pol.hymen_one.replicate, 
              corr=corSpatial(form= ~coords.x1+coords.x2, type ="spherical"))

AIC(mod, mod.exp, mod.gau, mod.lin, mod.rat, mod.sp)
```

mod is the best model since it has the lowest AIC. This means that spatial variation as a random effect is not a significant predictor of the diveristy. This might be due to the reduction in significance due to the increased df. 

```{r}
summary(mod)
```
When I pull up the outputs of the model, human population is not a significant predictor of pollinator density. This can be represented graphically. 
```{r}
pol.hymen_one.replicate %>% 
  ggplot(aes(x = hum.pop, y = diversity)) + 
  geom_point()+
  geom_smooth(method = "lm")

pol.hymenoptera %>% 
  ggplot(aes( x = hum.pop, y = diversity, colour = family))+
  geom_point()+
  geom_smooth(method = "lm")+ 
  facet_wrap(~family)
```

lets test this per family
```{r}
#in a dataframe:
pol.hymen.df <- pol.hymenoptera%>% 
    dplyr::select(family, diversity, hum.pop) %>% 
    group_by(family) %>% 
    do(mod = lm(diversity~ hum.pop, data = .)) %>% 
    mutate(AIC.value = AIC(mod))

pol.hymen.df <- tidy(pol.hymen.df, mod, AIC.value + family) %>%
    dplyr::select(family, term, estimate, std.error, p.value, AIC.value) %>% 
    arrange(family, desc(p.value))
       
pol.hymen.df
summary(mod)

#report the call, sample size, degrees of freedom, f-stat, r^2, adjusted r^2, 
```
None are significant. 


Moran's I: tests if points closer together are more or less similar to eachother compared to points further away

Set up:
```{r}
#first i need to modify the coordinates so that they correspond to the one replicate per gps coordinate dataset
coord.pol.one <- SpatialPoints(cbind(pol.hymen_one.replicate$long, pol.hymen_one.replicate$lat),
                                  proj4string=CRS("+proj=longlat +ellps=WGS84"))

sinu.pol.one <- spTransform(coord.pol.one, CRS("+proj=sinu +ellps:7271"))

pol.hymen_one.replicate <- bind_cols(pol.hymen_one.replicate, as.data.frame(sinu.pol.one))

#then i need to creat the distance matrix
dist.matrix <- as.matrix(dist(data.frame(sinu.pol.one)))

inv.dist <- 1/dist.matrix

diag(inv.dist) <- 0

```

Response Variable:
```{r}
#rearrange dataframe: genus relative abundances per state

response <- Moran.I(pol.hymen_one.replicate$diversity, inv.dist, alternative="two.sided", na.rm = TRUE, scaled = TRUE) 
```

The P-value is significant. This means that sites closer together have more similar diveristy characteristics. 

Predictor Variable:
```{r}
predictor <- Moran.I(pol.hymen_one.replicate$hum.pop, inv.dist, alternative="two.sided", na.rm = TRUE, scaled = TRUE) 
```

The p-value for our predictor variable is also significant. This means that closer sites also have more similar human populaitons. 

# _Combining all questions_: all singificant results in an inclusive models

model_full was the best model from question1. human population and spatial variation is not added to the model since it was not significant in predicting diveristy in question 3. 







